Predicting Ensemble Width and Location in Binaural Recordings of Music with Convolutional Neural Networks

### 2024-06-10 paper/main.tex
Update paper: related studies

\begin{abstract}
  Binaural audio technology has been around for many years, but its popularity has dramatically increased over the past decade due to advancements in virtual reality and streaming technologies. Along with its growing popularity, the quantity of publicly accessible binaural audio materials has also expanded. Consequently, there is now a need for automated and objective measurements of spatial content information, with ensemble width and location being the most important. In this study, 30 head-related transfer functions and 192 binaural music recordings from publicly accessible multi-track recording repositories were used to synthesize 23,040 binaural recordings. The synthesized recordings were then used to train a convolutional neural network prediction model, the aim of which was to predict the width and location of the ensemble for unseen recordings. The results indicate that models of ensemble breadth and position can be successfully constructed with low prediction errors --- $4.76\degree$ ($\pm0.10\degree$) for ensemble location and $8.57\degree$ ($\pm0.19\degree$) for ensemble width. This approach is the first of its kind to predict both ensemble width and location, offering a more accurate representation of spatial properties. This suggests significant potential for advancing spatial audio applications in virtual reality and streaming technologies, by providing audio engineers with tools that can leverage these methods to enhance spatial audio experiences.
\end{abstract}

@@

The human auditory system demonstrates exceptional proficiency in segregating, localizing, and interpreting diverse auditory signals, despite being limited to two ears. This capability arises from the intricate analysis of temporal, amplitude, and spectral disparities, a process termed binaural hearing \cite{blauert_spatial_1996}. This enables precise localization of sound sources in complex auditory environments. A particularly useful feature of binaural hearing is the 'cocktail party effect', which exemplifies the system's ability to focus on specific sounds while suppressing background noise \cite{cherry_experiments_1953}. An understanding of the auditory system is essential for an enhanced comprehension of its limits, the creation of more immersive binaural experiences for entertainment purposes \cite{zhang_surround_2017}, and the improvement of hearing aid devices to enhance auditory signal reception and improve spatial awareness \cite{thiemann_speech_2016}.

@@

The findings demonstrate that this method is effective in accurately predicting the spatial characteristics of sound sources in near-real-world scenarios. One of the key advantages of the proposed method is that it does not assume the number of audio sources. However, a significant limitation of this study is the lack of reverberation in the synthesized recordings.

@@

The majority of existing literature on the subject of sound source localization employs techniques that leverage the advantages of microphone arrays with more than two channels \cite{kaveh_statistical_1986, pavlidi_real-time_2012, pan_multi-tone_2021, hahmann_sound_2022, chung_sound_2022, liu_sound_2022}. In the context of sound source localization in binaural signals, the predominant focus of research is on the identification of individual sound sources, rather than groups of sounds \cite{dietz_auditory_2011, may_probabilistic_2011, may_binaural_2012, woodruff_binaural_2012, may_robust_2015, ma16c_interspeech, ma_exploiting_2017, benaroya_binaural_2018}. In terms of source direction of arrival (DOA), the majority of research assumes a fixed number of sound sources \cite{pang_multitask_2019, vecchiotti19, ma_exploiting_2017, woodruff_binaural_2012, s_spatiogram_2021}, which limits its practical applications as this information is rarely known in real-life binaural recordings. Moreover, the majority of studies have focused on relatively homogeneous signals, namely speech \cite{dietz_auditory_2011, may_probabilistic_2011, may_binaural_2012, woodruff_binaural_2012, may_robust_2015, ma16c_interspeech, ma_exploiting_2017, benaroya_binaural_2018, wang_binaural_2020, liu_multiple_2018, yang_deepear_2022, ma_robust_2018}.

In contrast to the aforementioned studies, the proposed method does not have any limitation to the number of sources. Furthermore, the approach is not limited to the speech, and indeed has been employed to work with much more diverse musical datasets, comprising sound sources such as instruments and vocals. In contrast to previous studies that have focused on individual sources, the proposed method does not aim to separate them but rather considers them as an ensemble, or in this case, a musical ensemble. To the authors' knowledge, this is one of the first methods to localize ensemble width (see \cite{antoniuk2023blind} for the previous study) and the first to localize both ensemble position and width simultaneously using a multi-task model.

>>>>

\begin{abstract}
  Binaural audio technology has been around for many years, but its popularity has dramatically increased over the past decade due to advancements in virtual reality and streaming technologies. Along with its growing popularity, the quantity of publicly accessible binaural audio materials has also expanded. Consequently, there is now a need for automated and objective measurements of spatial content information, with ensemble width and location being the most important. This study presents a novel method for predicting ensemble width and location in binaural recordings. To this end, 30 head-related transfer functions and 192 binaural music recordings from publicly accessible multi-track recording repositories were used to synthesize 23,040 binaural recordings. The synthesized recordings were then used to train a multi-task convolutional neural network prediction model, the aim of which was to predict the width and location of the ensemble for unseen recordings. The results indicate that models of ensemble breadth and position can be successfully constructed with low prediction errors --- $4.76\degree$ ($\pm0.10\degree$) for ensemble location and $8.57\degree$ ($\pm0.19\degree$) for ensemble width. This approach is the first of its kind to predict both ensemble width and location, offering a more accurate representation of spatial properties. This suggests significant potential for advancing spatial audio applications in virtual reality and streaming technologies, by providing audio engineers with tools that can leverage these methods to enhance spatial audio experiences.
\end{abstract}

@@

The human auditory system demonstrates exceptional proficiency in segregating, localizing, and interpreting diverse auditory signals, despite being limited to two ears. This capability arises from the intricate analysis of temporal, amplitude, and spectral disparities, a process termed binaural hearing \cite{blauert_spatial_1996}, which enables precise localization of sound sources in complex auditory environments. An important advantage of binaural hearing is demonstrated by the `cocktail party effect', which showcases the system's capacity to concentrate on individual sounds while suppressing background noise \cite{cherry_experiments_1953}. An understanding of the auditory system is essential for comprehending its limits and creating more immersive binaural experiences for entertainment purposes \cite{zhang_surround_2017}. It also helps to improve hearing aid devices by enhancing auditory signal reception and improving spatial awareness \cite{thiemann_speech_2016}.

@@

The findings demonstrate that this method is effective in accurately predicting the spatial characteristics of sound sources in near-real-world scenarios. Furthermore, this paper presents an experiment framework that allows for the objective measurement of a binaural localization technique in an objective manner, utilizing a large-scale dataset synthesized from real-world examples of music signals (for other use cases for this framework, see \cite{antoniuk2023blind, zielinski_automatic_2022, zielinski_spatial_2022, zielinski_comparison_2020}). One of the key advantages of the proposed method is that it does not assume the number of audio sources. However, a significant limitation of this study is the lack of reverberation in the synthesized recordings, which represents a material for future research.

@@

The majority of existing literature on the subject of sound source localization employs techniques that leverage the advantages of microphone arrays with more than two channels \cite{kaveh_statistical_1986, pavlidi_real-time_2012, pan_multi-tone_2021, hahmann_sound_2022, chung_sound_2022, liu_sound_2022}. In the context of sound source localization in binaural signals, the predominant focus of research is on the identification of individual sound sources, rather than groups of sounds \cite{dietz_auditory_2011, may_probabilistic_2011, may_binaural_2012, woodruff_binaural_2012, may_robust_2015, ma16c_interspeech, ma_exploiting_2017, benaroya_binaural_2018}. In terms of source direction of arrival (DOA) methods, the majority of research assumes a fixed number of sound sources \cite{pang_multitask_2019, vecchiotti19, ma_exploiting_2017, woodruff_binaural_2012, s_spatiogram_2021}, which limits its practical applications as this information is rarely known in real-life binaural recordings. Moreover, the majority of studies have focused on relatively homogeneous signals, namely speech \cite{dietz_auditory_2011, may_probabilistic_2011, may_binaural_2012, woodruff_binaural_2012, may_robust_2015, ma16c_interspeech, ma_exploiting_2017, benaroya_binaural_2018, wang_binaural_2020, liu_multiple_2018, yang_deepear_2022, ma_robust_2018}.

In contrast to the aforementioned studies, the proposed method is not limited by the number of sources. Furthermore, the approach is not limited to speech and has been applied to much more diverse musical datasets, including sound sources such as instruments and vocals. In contrast to previous studies that have focused on individual sources, the proposed method does not aim to separate them, but rather considers them as an ensemble, or in this case, a musical ensemble. This approach is inspired by the way real musical ensembles are arranged on stage. To the authors' knowledge, this is one of the first methods to localize ensemble width (see \cite{antoniuk2023blind} for the previous ensemble-width-related study), and the first to localize both ensemble position and width simultaneously using a multi-task model.

Sound localization methods can be classified into two primary categories based on the implementation of their underlying algorithms: glass-box and black-box. The glass-box methods, which are more traditional in the literature, rely on manual designing algorithms that mimic the auditory system to explicitly extract key features from the signal for location prediction, such as one of combination of the following: interaural level difference, interaural time difference, interaural coherence or interaural phase difference (see \cite{blauert_spatial_1996} for features description). In contrast, black-box methods typically employ only basic feature engineering and rely heavily on modern machine learning techniques to both extract features and make the final prediction. Consequently, they are less reliant on manual intervention (see \cite{vecchiotti19} for an example). The black-box algorithm extracts features implicitly and internally, without knowledge of their internal purpose or the equivalent of the human auditory system. This renders such methods less transparent and less predictable. This, in conjunction with the fact that they are typically implemented with deep neural networks with a large number of learning parameters, necessitates the development of them with the use of very large datasets, typically comprising hundreds of thousands of examples.
